---
title: "Practical Machine Learning Prediction Assignment"
author: "William Hogan"
date: "November 12, 2016"
output: html_document
---

## Introduction
This paper discusses how I created a model to characterize activity based on wearable data.
All of the models in this document are created via the caret package.

```{r}
library(caret)
```


## Data Aquisition and Preprocessing
There are initial two sets of data:  
* training  
* testing  

Here I download the data sets:
```{r}
### Import the data
pml_training <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
pml_testing <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(pml_training, destfile = "pml_training.csv")
download.file(pml_testing, destfile = "pml_testing.csv")
training <- read.csv("pml_training.csv", stringsAsFactors=FALSE, na.strings = c("", "#DIV/0", "NA"))
testing <- read.csv("pml_testing.csv", stringsAsFactors=FALSE, na.strings = c("", "#DIV/0", "NA"))

```

With the data on hand, there are two issues with the training data:  
* NA's throughout  
* first seven columns are arbitrary  

Here, I clean the data by:  
* Remove the columns with NA's  
* Remove columns 1:7  
* Check for variables with Near Zero Variance in the remaining data  
    + There were none; none remained

```{r}
### Clean the data
training_clean <- training[, colSums(is.na(training)) == 0]
testing_clean <- testing[, colSums(is.na(training)) == 0]

### Remove columns 1-7
training_clean <- training_clean[-c(1:7)]
testing_clean <- testing_clean[-c(1:7)]


### Check for near zero variance
nearZeroVar(training_clean)
```

###  Model Consideration and Testing
In selecting a model, I considered three different models:  
* rpart  
* k nearest neighbors  
* Random Forest

To test the model validity I partitioned the training data into a train and test set.
For each model, the test set was used to judge the accuracy.

```{r}
### Partition data

set.seed(3456)
inTrain <- createDataPartition(training_clean$classe, p=0.7, list=FALSE)
training_clean_train <- training_clean[inTrain, ]
training_clean_test <- training_clean[-inTrain, ]
```

For each model considered, a model is created using the training_clean_train data and then tested with the training_clean_test data.

The accuracy of the three models differed significantly:

Model Accuracy  
rpart 49.1%  
knn   90.8%  
rf    99.4%  

Of course for the prediction portion of the assignment, I will select the RF model.

By calling the final model of the rf model we get the OOB estimate for the error rate:
OOB estimate of  error rate: 0.71%

The models were cross validated via the data partitioning, and also as part of the optimization and internal selection that occurs within the caret package.


Here I create the three models.
After the RF model is completed, I call the finalModel to determine the OOB error rate.

```{r}
### rpart
m_rpart <- train(classe ~ ., data = training_clean_train, method = "rpart")
m_rpart
m_rpart$finalModel
test_output <- predict(m_rpart, training_clean_test[, -53])
mean(training_clean_test[, 53] == test_output)
table(training_clean_test[, 53], test_output)


### knn
m_knn <- train(classe ~ ., data = training_clean_train, method = "knn")
m_knn
m_knn$finalModel
test_output <- predict(m_knn, training_clean_test[, -53])
mean(training_clean_test[, 53] == test_output)
table(training_clean_test[, 53], test_output)

### rf
m_rf <- train(classe ~ ., data = training_clean_train, method = "rf")
m_rf
m_rf$finalModel
test_output <- predict(m_rf, training_clean_test[, -53])
mean(training_clean_test[, 53] == test_output)
table(training_clean_test[, 53], test_output)

```


Here, I call the final model to see what are the most important variables in the RF model.

```{r}
tmp_2 <- m_rf$finalModel
rfImp <- varImp(m_rf, scale = FALSE)
plot(rfImp, top = 20)

```


Here, I create a data frame to compile the predictions of the three models to compare their output for the final 20 row Test set.

```{r}
N <- 20
compiled_results <- data.frame(
  id = numeric(N),
  rpart = character(N),
  knn = character(N),
  rf = character(N))

compiled_results$id <- 1:20
test_output_final <- predict(m_rpart, testing_clean[, -53])
compiled_results$rpart <- test_output_final
test_output_final <- predict(m_knn, testing_clean[, -53])
compiled_results$knn <- test_output_final
test_output_final <- predict(m_rf, testing_clean[, -53])
compiled_results$rf <- test_output_final

compiled_results
write.table(compiled_results, "compiled_results.csv")

```

